---
title: "Data Exploration Report"
author: "Shafayet Shariar Hossain"
date: "4/30/2022"
output: html_document
---

## Load packages and data

```{r setup, message = FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)
library(svglite)
library(knitr)

clean_data <- read.table(here::here('Data','SEER_breast_cancer_Clean.csv'), header = TRUE, sep = ',')

## Change Variable Types

clean_data <- clean_data |>
  mutate_at(vars(-Year.of.diagnosis,-Survival.months),as.factor)

```

## Data Exploration

The SEER breast cancer data involves patient demographic information i.e. age, sex, race and their diagnostic information regarding breast cancer. The age variable in the dataset is grouped data that is why exact age of the patient is unknown. The dataset also includes variables regarding the surgery treatment of the patients and their corresponding survival months. Therefore, exploratory data analysis can reveal questions such as age distributions of breast cancer patients and many more. Below some exploratory analysis have been done to extract some insight from the dataset.

## Age Distribution of Breast Cancer Patients


```{r}

age_distribution <- ggplot(clean_data |>
                          group_by(Age_group)|>
                          summarise( N = n()))+
  aes(x=Age_group, y=N) + 
  geom_bar(stat = "identity") +
  coord_flip()+
  labs(title = 'Age distribution of Breast Cancer Patients')

age_distribution
```


The age distribution of the breast cancer patients shows that most of the patients are aged between approximately 50-70 years. The distribution peaks mostly around those groups. However, there are also significant amount of breast cancer patients that are relatively young. The distribution also resembles normal distribution. In the lower age groups, there are very few patients in number. That is why those bars are not visible in the graph. 

## Age group by Race


```{r, message=FALSE}

agexrace <- ggplot(clean_data |>
                            group_by(Age_group,Race)|>
                            summarise( N = n()))+
  aes(x=Age_group, y=N, fill = Race) + 
  geom_bar(stat = "identity") +
  labs(title = 'Barplot of Age by Race of Breast Cancer Patients')

agexrace <- agexrace + guides(x = guide_axis(angle = 45))

agexrace


```


From the plot it can be observed that in each of the age groups most of the breast cancer patients are predominantly white and the second most frequent race is black. Therefore, it provides an additional insight that most breast cancer patients are white across all age groups in US.


## Survival Months distribution by Race

```{r, message=FALSE}
survivalxrace <- ggplot(clean_data)+
  aes(x=Survival.months, y = Race, fill = Race) + 
  ggridges::geom_density_ridges(size = .2, alpha = .5, scale = 3) +
  labs(title = 'Survival distribution by Race')

survivalxrace
```


The survival months distribution across different race shows all the race experience similar distribution in terms of their survival months. Therefore, we can gather insight that the survival time of patients haven't been affected by the race they are part of. 

## Survival Months Distribution by Cancer Grade

```{r, message=FALSE}
survivalxgrade <- ggplot(clean_data|>
                           mutate(Cancer_grade = fct_relevel(Cancer_grade,
                                                             "Grade I", "Grade II",
                                                             "Grade III", "Grade IV")))+
  aes(x=Survival.months, y = Cancer_grade, fill = Cancer_grade) + 
  ggridges::geom_density_ridges(size = .2, alpha = .5, scale = 3) +
  labs(title = 'Survival distribution by Cancer_grade')

survivalxgrade
```

The stage of a cancer describes the size of a tumour and how far it has spread from where it originated. It describes the appearance of the cancerous cells. The staging of cancer is done through cancer grade and the higher the grade the higher the spread of the cancerous cells. The above graph shows the distribution of survival months in different cancer grades. And it can be observed that grade IV has a completely different survival month distribution compared to other cancer grades. Also, grades I II and III have similar distribution. 

## Survival Months by Age Group

```{r, message=FALSE}
survivalxage <- ggplot(clean_data) +
  aes(x = Age_group,
      y = Survival.months,
      color = Age_group) +
  geom_boxplot()+
  labs(title = "Boxplot of Survival Months by Age Group")


survivalxage <- survivalxage + guides(x = guide_axis(angle = 45))

survivalxage
```


The above graph shows the quantiles of survival months across different age groups. The boxplot shows that the median survival months are almost same across all age groups and provides insight that higher age or lower age doesn't have any impact on survival of the breast cancer patients. 


## Survival Months by Primary Site

```{r, message=FALSE}
survivalxsite <- ggplot(clean_data) +
  aes(x = Primary_site,
      y = Survival.months,
      color = Primary_site) +
  geom_boxplot()+
  labs(title = "Boxplot of Survival Months by Primary Site")


survivalxsite <- survivalxsite + guides(x = guide_axis(angle = 45))

survivalxsite
```



The above graph shows whether the quantiles of the survival months of the breast cancer patients different across primary site of the cancer. Primary site of the cancer means where the cancerous origininated initially. In the graph the survival months haven't differed across different primary sites of the cancer. 

## Survival Months by surgery


```{r, message=FALSE}
survivalxsurgery <- ggplot(clean_data) +
  aes(x = Surgery,
      y = Survival.months,
      color = Surgery) +
  geom_boxplot()+
  labs(title = "Boxplot of Survival Months by Surgery")


survivalxsurgery <- survivalxsurgery + guides(x = guide_axis(angle = 45))

survivalxsurgery
```


The above graph shows the distribution of survival months for patients who had surgery and who hadn't. From the boxplot it can be concluded that the median survival time is higher for patients who had surgery compared to those who did not have a surgery. Therefore, there is a hint that surgery has helped prolonging patient's survival. 

## Survival months by Laterality

```{r, message=FALSE}


survivalxlaterality <- ggplot(clean_data) +
  aes(x = Laterality,
      y = Survival.months,
      color = Laterality) +
  geom_boxplot()+
  labs(title = "Boxplots of Survival Months by Laterality")


survivalxlaterality <- survivalxlaterality + guides(x = guide_axis(angle = 45))

survivalxlaterality
```


Cancer laterality of breast cancer means which side of the breasts the cancerous cells has spread. The above graph shows the distribution of survival months across the laterality of breast cancer. From the graph it can be observed that patients with cancer only one side of the breast cancer (left/right) has higher median survival month compared to other lateralities. It also makes sense as cancer cells on one side of the pair of breast will have higher survival times compared to both. 

## Survival Months by Cause of Death


```{r, message=FALSE}

survivalxcod <- ggplot(clean_data) +
  aes(x = Cause_of_death,
      y = Survival.months,
      color = Cause_of_death) +
  geom_boxplot()+
  labs(title = "Boxplot of Survival Months by Cause of Death")


survivalxcod <- survivalxcod + guides(x = guide_axis(angle = 45))

survivalxcod

```


The above graph shows the survival month distribution of patients who died because of breast cancer and who died of other causes. The median survival month of patients who died of breast cancer is lower compared to those who are alive or dead because of other disease.

## Proportions of Surgery on Cancer Histology type.


The cancer histology type is a description of a tumor based on how abnormal the cancer cells and tissue look under a microscope and how quickly the cancer cells are likely to grow and spread. We wanted to see if there is any histologic type which didn't require surgery rather other treatment procedures e.g. chemotherapy, radiationtherapy etc. The following table shows the proportion of patients who had surgery and who hadn't based on their corresponding histology type. The proportions tells us in each of the histology types majority of the patients had surgery and no histology type had an opposite case where the majority did not have a surgery. 

```{r, message=FALSE}
clean_data |>
  group_by(Histology_Type, Surgery) |>
  summarise(n=n()) |>
  mutate(prop = n/sum(n)) |>
  subset(select = c("Surgery", "Histology_Type", "prop"))|>
  spread(Surgery, prop) |>
  kable()
```


## Conclusion

In summary, from the above exploratory analysis we have learned that most of the breast cancer patients are aged people with most of them being white.There are no differences in survival distribution across different races in US population. The survival time of patients differ if the patient is in Grade IV cancer level. Other than that the survival distribution doesn't differ across different age groups and primary site of the cancer. However, patients who had surgery had survived longer than those who did not and patients with cancer one side of the breasts also have higher survival times. In terms of which type of patients require surgery there haven't been any finding that certain histology type doesn't require surgery. 

